---
title: "Tutorial 7: Logistic Regression and Time Series Forecasting"
author: 'Ali'
date: 'Due by October 17, 9:00 AM'
output: html_document
---

## Preparation

```{r load-libraries, echo=FALSE, warning = FALSE, message = FALSE}
# load required packages
# install any package below if it's first time loaded in your computer.
library(dplyr)
library(tidyr)
library(tseries) 
library(TTR) # One alternative for time-series in R
library(forecast) # An alternative for time series in R
library(car) # "Companion to Applied Regression" package, for F-test for linear combination of regression coefs
library(wooldridge) # wooldridge data set will be used in this tutorial
library(ggplot2) # optional. we expect you to know base graphics, but allow ggplot if you find it easier
```

## Part Two: Assignment Submission (25 points)

### Question 3 (total 17 points)

We will study fertility rates with two time-series datasets on U.S. and Singapore, respectively. 

High fertility rate is essential for long-term growth in any economy. Many countries are troubled with low or even negative fertility rate. For example, the fertility rate in Singapore in general shows a decreasing trend in recent decades (https://www.channelnewsasia.com/news/singapore/number-of-babies-born-in-singapore-falls-to-lowest-in-8-years-11743722). Fertility is affected by many socio-economic factors, including single rate, family disposable income, level of tax duty, war attrition, contraception technology, etc. 

Let's first take a look at the fertility rate in the United States in 20th century. `fertil3` data contains information about woman's fertility rate and personal tax exemption in U.S. in early-mid 1900s. 

- Dataset required: `data('fertil3')` in `wooldridge` package.

(Note: This dataset comes from a publicly available dataset from Jeffery Wooldridge Textbook. See data description in https://rdrr.io/cran/wooldridge/man/fertil3.html)

First, load in the time series data for this question. There are 72 observations on 24 variables about women fertility rate between year 1913 and 1984. Key variables are listed below:

- `gfr`: births per 1000 women between age 15 and 44.
- `pe`: real value personal tax exemption in US dollars.
- `pe_1`: 1-period lag of `pe`, real value personal tax exemption in US dollars from last year.
- `pe_2`: 2-period lag of `pe`, real value personal tax exemption in US dollars from two years ago.
- `ww2`: a binary variable = 1 during World War 2 between 1941 and 1945.
- `pill`: a binary variable = 1 from 1963 on when the birth control pill was made available for contraception.

```{r q3-read-dataset, echo=TRUE}
# read dataset into workplace, note that you need library(wooldridge) to load this data set
data('fertil3')
# if you want, is can be converted to ts object
fertil = ts(fertil3, frequency = 1, start = 1913)
```

**Q3(a) Start off by plotting `gfr` (fertility rate) and `pe_1` (personal tax exemption from last year) against time in the same plot. What do you observe from the time series plots alone: do you see any trend or seasonality? Do you think that `gfr` time series is stationary? (2 points)**

Remark: You should be able to visualize data with R base graphics at least. Other alternative R graphic packages such as `ggplot` are welcomed additions.

```{r q3a, echo=TRUE}
plot(fertil3$year, fertil3$gfr, main = "Fertility Rate & Personal Tax Exemption from Last Year Against Time", ylab = "Fertility and Personal Tax Exemption", xlab = "Year", type = "l", ylim = c(0,250), col = "purple")
lines(fertil3$year, fertil3$pe_1, col = "brown")
legend("bottomright", legend=c("Personal Tax Exemption", "Fertility Rate"), col=c("purple", "brown"), lty = 1:1)

```

Downward trend for fertility rate & no seasonality since the graph does not repeat at certain lengths of intervals. Time series is non-stationary. <br>

Upward trend, followed by a spike in personal tax exemption in between the years 1940 and 1950, followed by a downward trend. There is no seasonality since the graph does not repeat at certain lengths of intervals.

**Q3(b) Many labor economists believe that fertility rate is affected by the economic policy such as personal tax exemption, e.g., `pe` and their lag terms. Run a linear regression of `gfr` on `pe`, `pe_1`, and `pe_2`. Interpret the coefficient before `pe`. Is it statistically significant? What's your conclusion about whether tax exemption improves fertility rate, from the regression output? (3 points)**

```{r q3b, echo=TRUE}
pe = lm(gfr~pe + pe_1 + pe_2, data= fertil3)
summary(pe)
```

As p-value (intercept) is less than 0.05, we reject the null hypothesis that β = 0. Hence there is a significant relationship between the variables in the linear regression model of the data set, and the intercept is significant. The average fertility rate is 93.2 (3s.f) births per 1000 women, as observed from the intercept value, if the personal tax exemption is 0 US dollar. <br>

Slope for pe_1 is -0.02134. When personal tax exemption increases by one unit in each year, there should be an average decrease in fertility rates. <br>

Despite this, the slope for pe_2 is 0.0539. When personal tax exemption increases by one unit in two years, there is an average increase in fertility rates. In conclusion, personal tax exemption may improve the fertility rate in the short run (pe_1 is a 1-year lag), but lowering fertility in the long run (pe_2 is a 2-year lag).


**Q3(c) Now, include a time trend variable `t` into the regression model in part (b). Interpret the coefficients for `pe_2` and `t`. What's been changed in the regression result compared to previous one, in terms of coefficients (its sign and magnitude), goodness-of-fit R-square and F-test of the linear regression model? Which model do you think we should choose, and why? (4 points)**

```{r q3c, echo=TRUE}
pe_t = lm(gfr~pe + pe_1 + pe_2 + t, data = fertil3)
summary(pe_t)
```

t has a slope of -1.07 (3s.f). Fertility rate falls as time passes. pe_2 and goodness of fit R square increased in magnitude, while the p-value of F-test decreased in magnitude significantly (still positive in sign). As the p-value is now 3.642e-13, the t model provides sufficient evidence to conclude that the variance between the means of two populations are significantly different. Given this, the new model should be chosen.

As we have discussed fertility rate in U.S. between 1913 and 1984 in previous question, now let's turn our attention to fertility rate in Singapore in modern times, where the fertility rate plummets to lowest figure in 2020 in history. 

- Dataset required: `SGfertil20.csv`

Note: This dataset comes from publicly available data from the Singapore Department of Statistics, or SingStat. https://www.singstat.gov.sg/find-data/search-by-theme/population/births-and-fertility/latest-data. You can download the data yourself and explore. For homework, I have done some data cleaning as to have only one time series variable ,`gfr`, in `sgfertil.csv` which contains information on fertility rate of childbearing age women between age 15 and 44 in Singapore comparable to that in U.S. 

```{r q3-fertil-data-input, echo=TRUE}
# read raw data from 'sgfertil.csv'
sgfertil = read.csv(file = 'SGfertil20.csv', header= TRUE)
colnames(sgfertil)[1]='Data.Series'

sgfertil = sgfertil %>% 
  # given our focus in univariate time-series analysis on total fertility rate..
  select(Data.Series, Total.Fertility.Rate..Per.Female.) %>%
  # rename
  rename(year = Data.Series, gfr = Total.Fertility.Rate..Per.Female.) %>%
  # sort the data bt ascending on year
  arrange(year)
# extract the key time series of our interest and convert it to ts obj
gfrsg = ts(sgfertil$gfr, start = 1960, end = 2020, frequency = 1)
```
`sgfertil` contains yearly data about gross fertility rate per thousand women between 1960 and 2020. A quick description of the data:

Obs:  61
- `gfr`: total fertility rate of women in their childbearing age between 15 and 44, #births per thousand
- `year`: as described.

**Q3(d) First plot the `gfrsg` against time. Draw the time series plot of `gfrsg`. Describe what you observe from the plot: do you see any trend or/and seasonality? Do you think total fertility rate time series in Singapore `gfrsg` stationary? (2 points)**

```{r q3d, echo=TRUE}
plot(gfrsg, main = "Total Fertility Rate Against Time", xlab = "Year", ylab= "Total Fertility Rate", type = "l", col = "plum")
```
Downward trend in fertility rate against time in Singapore. No seasonality since the graph does not repeat at certain lengths of intervals. This time series is non-stationary as it trends downwards.

**Q3(e) In order to predict future fertility rate using information contained in the time series of `gfrsg` itself, let's make forecast by simply fitting a moving average series. First, compute the moving averages with window length of 4 years and plot both the original and moving average series in the same graph. Second, obtain the moving average prediction series. What is the MA predicted value for Year 2021? (3 points)**

```{r q3e, echo=TRUE}
plot(gfrsg, main="Original Series and Moving Average Series against Time")
lines(TTR::SMA(gfrsg, n=4), col = "red")
legend("topright", cex = 1, col = c("black", "red"), lty = 1, c("Original Series", "Moving Average Series"))

prediction_2021 <- TTR::SMA(gfrsg, n=4)[61]
prediction_2021
```
MA predicted value for year 2021 is 1.135.

**Q3(f) Based on your observation in part (d), predict the fertility rates for Year 2021 and Year 2024 with Holt-Winter model of your choice. (3 points)**

```{r q4a, echo=TRUE}
hw <- HoltWinters(gfrsg, gamma = FALSE)
hw

hw_prediction2021 <- predict(hw, n.ahead = 1)
hw_prediction2024 <- predict(hw, n.ahead = 4)

hw_prediction2021
plot(hw, hw_prediction2021, main = "2021 Holt-Winters Filtering")

hw_prediction2024
plot(hw, hw_prediction2024, main = "2024 Holt-Winters Filtering")
```

2021's predicted fertility = 1.08 (3s.f)
2022's predicted fertility = 1.02 (3s.f)

### Question 4 (Total 8 points)

- Dataset required: `data('recid')`

Recidivism rate in Singapore is 24% in 2016 (https://data.gov.sg/dataset/recidivism-rate). Criminals tend to relapse into criminal offense after the release from the prison. Recidivism is costly and causes serious social and economical problem. It is not only wasteful with the resources invested in prison, including staffing, infrastructure investment, daily operation cost, and economic opportunity cost for both prisoners and staffs (i.e., labor values that could be generated elsewhere other than being locked up and guarding in prison, respectively), but also harms the community for the second time due to crime re-commitment. Recidivism is thus a critical evaluating metric for prison performance, e.g., rehabilitation or training program. `recid.csv` contains 1445 observations of recidivism cases in United States where it currently has the largest prison population in the world (about one out every five peope imprisoned in the world is incarcerated in the United States).

```{r q3-data-input, echo=TRUE}
recid = read.csv(file = 'recid.csv', header= TRUE)
```

C.-F. Chung, P. Schmidt, and A.D. Witte (1991), “Survival Analysis: A Survey,” Journal of Quantitative Criminology 7, 59-98.
Data Source: https://www.cengage.com/cgi-wadsworth/course_products_wp.pl?fid=M20b&product_isbn_issn=9781111531041. A quick description of the data:

Obs:  1445

  1. black                    =1 if black
  2. alcohol                  =1 if alcohol problems
  3. drugs                    =1 if drug history
  4. super                    =1 if release supervised
  5. married                  =1 if married when incarc.
  6. felon                    =1 if felony sentence
  7. workprg                  =1 if in N.C. pris. work prg.
  8. property                 =1 if property crime
  9. person                   =1 if crime against person
 10. priors                   # prior convictions
 11. educ                     years of schooling
 12. rules                    # rules violations in prison
 13. age                      in months
 14. tserved                  time served, rounded to months
 15. follow                   length follow period, months
 16. durat                    max(time until return, follow) in month
 17. cens                     =1 if duration right censored
 18. ldurat                   log(durat)

**Q4(a) Criminal offense lawyers usually fight for supervised release of the offender. Given seven variables in `recid` data set: `super`, `rules`, `age`, `tserved`, `married`, `prior` and `black`, how could you help the law firm to predict the outcome of a recent case given that a 55-year-old married black client has been serving the jail time for 5 years and 11 months with no prior conviction and during which broken no rule in the prison?** 
**In order to evaluate the likelihood to answer such question. First of all, write down the regression model you propose to use. Clearly define or label your variables.  (2 point)**

Multivariate regression: There are more than 2 variables involved, `super` (release supervision), `rules` (number of rules violations in prison), `age` (age of observed person in months), `tserved` (time served, rounded to months), `married` (marriage status), `priors` (number of prior convictions) and `black` (race of the observed person).

**Q4(b) Run your regression model in R and interpret the coefficients before `priors` and `black`. Based on the results, do you think there was any racial discrimination in terms of likelihood of supervised release, and why? (3 point)**

```{r q4b, echo=TRUE}
rm <- lm(super ~ rules + age + tserved + married + priors + black, recid)
summary(rm)
```

`priors` slope = -0.0198 (3s.f), which is negative. One unit increase in the number of prior convictions leads to a decrease in supervised release (other independent variables are held constant). <br>

`black` slope = 0.0258 (3s.f), which is positive. When a person is black, there is a increase in supervised release (other independent variables are held constant). <br>

As such, there is no racial discrimination in terms of likelihood of supervised release.

**Q4(c) Finally, what is the predicted probability for supervised release for the new client? (2 point)**

```{r q4c, echo=TRUE}
newclient = data.frame(rules = 0, age = 55, tserved = 71, married = 1, priors = 0, black = 1)
prediction_newclient = predict(rm,newclient)
prediction_newclient
```
Predicted probability is 1.17. Supervised release is likely for the client.


**Q4(d) Successfully debug the Rmarkdown file and produce an HTML for submission. (1 point)**
