---
title: "Tutorial 3"
output: 
  html_document: default
  pdf_document: default
---

## Tutorial 3

### Preparation

#```{css, echo=FALSE}

#.rpivotTable{ overflow-x: scroll; }
#.rpivotTable{ overflow: auto; }

#```

```{r load-libraries, echo=TRUE}
#install.packages - only need to run this code once to install the package
# load required packages
# library("xxxx")
library("tidyverse") #need to call the library before you use the packages
library("rpivotTable")
library("knitr")
```

### Tutorial 3 Part 2 (To be submitted)

- Dataset required: `ds_salaries.csv`

The dataset contains Data Science jobs salary data aggregated by ai-jobs.net. There are 607 observations and 11 variables. Each of the variable (column) is defined as follows: 

- `work_year`: The year the salary was paid.
- `experience_level`: The experience level in the job during the year with the following possible values: EN Entry-level / Junior; MI Mid-level / Intermediate; SE Senior-level / Expert; EX Executive-level / Director
- `employment_type`: The type of employment for the role: PT Part-time; FT Full-time; CT Contract; FL Freelance
- `job_title`: The role worked in during the year.
- `salary`: The total gross salary amount paid.
- `salary_currency`: The currency of the salary paid as an ISO 4217 currency code.
- `salary_in_usd`: The salary in USD (FX rate divided by avgerage USD rate for the respective year via fxdata.foorilla.com).
- `employee_residence`: Employee's primary country of residence during the work year as an ISO 3166 country code.
- `remote_ratio`: The overall amount of work done remotely, possible values are as follows: 0 No remote work (less than 20%); 50 Partially remote; 100 Fully remote (more than 80%)
- `company_location`: The country of the employer's main office or contracting branch as an ISO 3166 country code.
- `company_size`: The average number of people that worked for the company during the year: S less than 50 employees (small); M 50 to 250 employees (medium); L more than 250 employees (large)

```{r q2-read-dataset, echo=TRUE}
#import dataset into RStudio

library(readr)
DS <- read_csv("C:/Users/aliya/OneDrive/Desktop/ds_salaries.csv")
View(DS)

```

As a business analytics analyst, you are required to perform the following tasks as described in each question below. 

### Q2.(ai) Job Information Dashboard Part 1 (4 marks)
- First, analyse the data to see how many job titles are captured in the dataset? Write you answer below.
- Next, create a table and barplot to visualise the frequency distribution for the top 10 Job titles. 

```{r q2.(ai), echo=TRUE}
Job_Titles <- DS %>%
  group_by(`job_title`) %>%
  summarise(n=n()) %>%
  arrange(desc(n))

# From the table above, there are 50 jobs

Top10Jobs <- Job_Titles %>%
  slice(1:10)
kable(Top10Jobs, caption = "Frequency of Job Titles")

barplot(Top10Jobs$n, main="Frequency Distribution of Top 10 Job Titles", xlab="Job Titles", ylab="Frequency", names.arg = Top10Jobs$job_title, cex.names = 0.2, col = "magenta")

```


### Q2.(aii) Job Information Dashboard Part 2 (6 marks)
- Create a table and pie chart for each of the following variables: `employment_type`, `experience_level`, `work_year` (6 marks)
```{r q2.(aii), echo=TRUE}
# employment_type
EmploymentType <- DS %>%
  group_by(`employment_type`) %>%
  summarise(n=n())
kable(EmploymentType, caption = "Frequency of Employment Types")

slice.employmenttype <- EmploymentType$n
employmenttype.piepercent <- 100*round(EmploymentType$n/sum(EmploymentType$n), 2)
label <- EmploymentType$employment_type
label <- paste(label, ",", sep=" ")
label <- paste(label, employmenttype.piepercent)
label <- paste(label, "%", sep=" ")
pie(slice.employmenttype,
    labels=label,
    col=c("blue", "cyan", "dodgerblue", "purple"),
    radius=1,
    main="Employment Type",
    cex = 0.5)
legend("bottomleft", EmploymentType$employment_type, cex = 0.8, fill = c("blue", "cyan", "dodgerblue", "purple"))

# experience_level
EL <- DS %>%
  group_by(`experience_level`) %>%
  summarise(n=n())
kable(EL, caption = "Frequency of Experience Level")

slice.el <- EL$n
el.piepercent <- 100*round(EL$n/sum(EL$n), 2)
label <- EL$experience_level
label <- paste(label, ",", sep=" ")
label <- paste(label, el.piepercent)
label <- paste(label, "%", sep=" ")
pie(slice.el,
    labels=label,
    col=c("red", "orange", "yellow", "green"),
    radius=1,
    main="Experience Level",
    cex = 0.75)
legend("bottomleft", EL$experience_level, cex = 0.8, fill = c("red", "orange", "yellow", "green"))

# work_year
WY <- DS %>%
  group_by(`work_year`) %>%
  summarise(n=n())
kable(WY, caption = "Frequency of Work Year")

slice.wy <- WY$n
wy.piepercent <- 100*round(WY$n/sum(WY$n), 2)
label <- WY$work_year
label <- paste(label, ",", sep=" ")
label <- paste(label, wy.piepercent)
label <- paste(label, "%", sep=" ")
pie(slice.wy,
    labels=label,
    col=c("green", "blue", "magenta"),
    radius=1,
    main="Work Year",
    cex = 0.75)
titles <- c("2020", "2021", "2022")
legend("bottomleft", titles, cex = 0.8, fill = c("green", "blue", "magenta"))

```

### Q2.(aiii) Job Information Dashboard Part 3 (5 marks)
- Create a table and chart to compare the frequency distributions for `employment_type` across the top 3 job titles.
- Create a table and chart to compare the frequency distributions for `experience_level` across the top 3 job titles.

- Based on charts in 2aii and 2aiii, describe briefly the type of jobs that are in the dataset. For example, is there more of a particular employment type in the data, and is the pattern the same for the top 3 job titles. 

```{r q2.(aiii), echo=TRUE}
# `employment_type` across the top 3 job titles
ETJT <- DS %>%
  select(job_title, employment_type) %>%
   filter(job_title == 'Data Scientist'| job_title == 'Data Engineer'| job_title == 'Data Analyst')
ETJT.table <- table(ETJT$employment_type, ETJT$job_title)

titles <- c("Data Analyst", "Data Engineer", "Data Scientist")
type <- c("FL", "FT", "PT")

barplot(ETJT.table, main = "Employment Type of Top 3 Job Titles", names.arg = titles, xlab = "Number of people", ylab = "Job Titles", col = c("Red", "Orange", "Yellow"), beside = TRUE, horiz = TRUE, cex.names = 0.8)
legend("bottomright", type, cex = 0.8, fill = c("Red", "Orange", "Yellow"))

# `experience_level` across the top 3 job titles
ELJT <- DS %>%
  select(job_title, experience_level) %>%
   filter(job_title == 'Data Scientist'| job_title == 'Data Engineer'| job_title == 'Data Analyst')
ELJT.table <- table(ELJT$experience_level, ELJT$job_title)

titles <- c("Data Analyst", "Data Engineer", "Data Scientist")
type <- c("EN", "EX", "MI", "SE")
colors <- c("green", "blue", "pink", "purple")

barplot(ELJT.table, main = "Experience Level of Top 3 Job Titles", names.arg = titles, xlab = "Number of people", ylab = "Job Titles", col = colors, beside = TRUE, horiz = TRUE, cex.names = 0.8)
legend("bottomright", type, cex = 0.8, fill = colors)

```

### Q2.(b) Salary Dashboard (10 marks)
This dashboard is to visualise the salary data captured in the dataset better. 

- i. Create a table and chart to visualise the frequency distributions for `salaryinusd`
- ii. Create separate tables and charts to visualise the frequency distributions for `salaryinusd` across the top 3 Job titles. Do you observe any difference across the 3 job titles? [HINT: you may want to use the same breaks/bin width for the 3 charts for ease of comparison]

```{r q2b, echo=TRUE}
#i
options(scipen = 100)
h.salary <- hist(DS$`salary_in_usd`,
                 main = "Histogram of Salary in USD",
                 xlab = "Salary in USD",
                 ylab = "No. of workers",
                 col=c("darkorange"),
                 ylim = c(0,350),
                 breaks = 8,
                 labels = TRUE)

salary.group <- cut(DS$`salary_in_usd`, h.salary$breaks, include.lowest = TRUE)
table.salary <- table(salary.group)
kable(table.salary, caption = "Frequency distribution by Salary in USD")

# ii
## Data Scientist
dataSci.salary <- DS %>%
  select(job_title, salary_in_usd) %>%
  filter(job_title == "Data Scientist")

h.dataSci.salary <- hist(dataSci.salary$`salary_in_usd`,
                         main = "Histogram of Data Scientists' Salary in USD",
                         xlab = "Salary in USD",
                         ylab = "No. of workers",
                         col=c("red"),
                         ylim = c(0, 80),
                         breaks = 4,
                         labels = TRUE)

salary.ofDataSci <- cut(dataSci.salary$`salary_in_usd`, h.dataSci.salary$breaks, include.lowest = TRUE)
t.dataSci.salary <- table(salary.ofDataSci)
kable(t.dataSci.salary, caption = "Frequency distribution of Data Scientists by Salary in USD")

## Data Engineer
dataEng.salary <- DS %>%
  select(job_title, salary_in_usd) %>%
  filter(job_title == "Data Engineer")

h.dataEng.salary <- hist(dataEng.salary$`salary_in_usd`,
                         main = "Histogram of Data Engineers' Salary in USD",
                         xlab = "Salary in USD",
                         ylab = "No. of workers",
                         col=c("orange"),
                         ylim = c(0, 80),
                         breaks = 4,
                         labels = TRUE)

salary.ofDataEng <- cut(dataEng.salary$`salary_in_usd`, h.dataEng.salary$breaks, include.lowest = TRUE)
t.dataEng.salary <- table(salary.ofDataEng)
kable(t.dataEng.salary, caption = "Frequency distribution of Data Engineers by Salary in USD")

### Data Analyst
dataAna.salary <- DS %>%
  select(job_title, salary_in_usd) %>%
  filter(job_title == "Data Analyst")

h.dataAna.salary <- hist(dataAna.salary$`salary_in_usd`,
                         main = "Histogram of Data Analysts' Salary in USD",
                         xlab = "Salary in USD",
                         ylab = "No. of workers",
                         col=c("red"),
                         ylim = c(0, 80),
                         breaks = 2,
                         labels = TRUE)

salary.ofDataAna <- cut(dataAna.salary$`salary_in_usd`, h.dataAna.salary$breaks, include.lowest = TRUE)
t.dataAna.salary <- table(salary.ofDataAna)
kable(t.dataAna.salary, caption = "Frequency distribution of Data Analysts by Salary in USD")

### Data scientists are the highest earning workers, followed by data engineers, then finally, data analysts.
```

### Q2.(c) Interactive Contingency Table (2 marks)
Create a contingency table with the `rPivotTable` function for Data Scientist jobs that displays frequency for the following 3 variables: experience_level, employment_type and work_year. 

```{r q2.(c), echo=TRUE}
con.table <- DS %>%
  select(experience_level, employment_type, work_year)

rpivotTable(con.table,rows=c("experience_level"), cols=c("employment_type", "work_year"))

```


### Q2.(d) Pareto Analysis (3 marks)
Conduct a Pareto analysis on `salary_in_usd` to be displayed in this dashboard. (Hint: check the number and percentage of individuals that contribute  to 80% of the total salary_in_usd earned in the dataset. Do you observe a 80-20 split? If not, what is the ratio?). 

```{r q2.(d), echo=TRUE}
salaries <- DS %>%
  select(salary_in_usd) %>%
  arrange(desc(salary_in_usd))

salaries$percentage <- salaries$salary_in_usd/sum(salaries$salary_in_usd)
salaries$cumulative <- cumsum(salaries$percentage)
salaries$cumulative <- as.numeric(rownames(salaries))/nrow(salaries)
which(salaries$cumulative > 0.8)
(which(salaries$cumulative > 0.8)) / nrow(salaries)

```